---
title: "Mini Data Analysis Milestone 2"
author: "Sasha Tuttle"
output: github_document
---

Fill in the sections that are commented out with `<!--- start your work here--->`.
When you are done, make sure to knit to an `.md` file by changing the output in 
the YAML header to `github_document`, before submitting a tagged release on canvas.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
Pilot<-read.csv("C:/Users/stutt/OneDrive/Desktop/Greenhouse_study/Pilot_study.csv", 
  header=T) 
suppressPackageStartupMessages(library(tidyverse))
```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. Do blueberry flowers of the bluecrop variety experience a greater fruit set when they receive pollen from another blueberry variety?

2. Do heavier berries have a higher TSS? a lower pH/TA?

3. Does stem diameter have an effect on blueberry weight?

4. Does the Bluecrop variety experience greater benefits from receiving pollen from another variety relative to the Duke variety?
<!----------------------------------------------------------------------------->

### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->


First, I need to clean up my data and separate the varieties like I did in Milestone 1.

```{r}
head(Pilot)

#Some of the columns aren't necessary as they were only used for calculations.
#Other columns were simply used for sample IDs or have poor choice of names.

Data<-as_tibble(Pilot) %>%
select(Variety, Pollen_donor, berry_weight, fruit_set,
Average_TSS, Average_pH, TA.true.) %>%
arrange(Pollen_donor, desc(fruit_set)) %>%
rename(Berry_weight_g = berry_weight) %>%
rename(TSS = Average_TSS) %>%
rename(pH = Average_pH) %>%
rename(Titratable_acid = TA.true.)


Blue_B<- as_tibble(Data) %>%
filter(Variety == "Bluecrop") %>%
filter(Pollen_donor =="Bluecrop")


Blue_D<- as_tibble(Data) %>%
filter(Variety == "Bluecrop") %>%
filter(Pollen_donor =="Duke") 


Blue_R<- as_tibble(Data) %>%
filter(Variety == "Bluecrop") %>%
filter(Pollen_donor =="Reka") 
```

Blue_B organizes the data into the self-pollinated Bluecrop treatment (meaning 
Bluecrop flowers that received pollen from another Bluecrop variety).

Blue_D organizes the data into Bluecrop variety flowers that received pollen 
from the Duke variety (an out-cross treatment).

Blue_R Blue_D organizes the data into Bluecrop variety flowers that received pollen 
from the Reka variety (an out-cross treatment).


**Exploring Research question 1**
**Summarizing the data (task 1)**
```{r}
summary(Data$fruit_set)

#alternatively, you can use the following functions:
mean(Data$fruit_set)
median(Data$fruit_set)
range(Data$fruit_set)
```

These functions have given me the summary statistics for all of the treatments though!
I am interested in comparing the mean and median between treatments.

```{r}
summary(Blue_B$fruit_set)
summary(Blue_D$fruit_set)
summary(Blue_R$fruit_set)
```

**Graph the summarized variables (Task 5)** 


```{r}
Blue<-as_tibble(Data) %>%
  filter(Variety == "Bluecrop")

ggplot(Blue, aes(Pollen_donor, fruit_set)) + geom_boxplot() + 
    geom_jitter(width=0.1, alpha=0.5) + xlab("pollen donor") + ylab("fruit set (%)") +
    ggtitle("Fruit Set vs Pollen Donor in the Bluecrop variety")
```

The box plot above illustrates the minimum value, median, maximum value, as well 
as the 1st and 3rd quartiles.The figures suggests that the self-crossed Bluecrop
treatments experienced a smaller fruit set relative to those that received pollen 
from another variety. I would have to use an anova to see if this difference is 
statistically significant though. 

**Exploring Research question 2**

I am interested to see if heavier blueberries have a higher % total soluble solids
(TSS) and/or a lower pH because these variables relate to blueberry quality. TSS correlates with sugar content in blueberries, so a large blueberry with a high TSS 
and low pH would be sweet and considered to be high quality. Alternatively, larger berries could have a lower TSS (%) as the sugar becomes diluted. I'm interested
to see if cross-pollination increases blueberry weight in Bluecrop but decreases
TSS as a trade off (affecting the sugar:acid ratio). But I hope to see an increase 
in overall blueberry quality as a result of increased genetic diversity (so, an 
increase in berry weight and TSS, and a decrease in pH and TA). 

**Summarizing the data**

It doesn't help me to find the summary statistics of the berry weight or
the other variables so instead I'll create a tibble to see if I can see any 
trends when organizing the data. 

```{r}
Data1<-as_tibble(Pilot) %>%
filter(Variety == "Bluecrop") %>%
select(Variety, Pollen_donor, berry_weight, Average_TSS, Average_pH, TA.true.) %>%
arrange(Variety, Pollen_donor, desc(berry_weight), Average_TSS) %>%
rename(Berry_weight_g = berry_weight) %>%
rename(TSS = Average_TSS) %>%
rename(pH = Average_pH) %>%
rename(Titratable_acid = TA.true.)

head(Data1, n=36)
```

Organizing the data like so, it does seem like heavier berries may have a higher TSS
in the control treatment (Bluecrop x Bluecrop) but the trend isn't as apparent in
the other treatments. Potential trends in the pH aren't as apparent either. I'll
create one scatter plot of berry weight against TSS and one with berry weight
against pH to see if these trends follow a positive linear relationship. 

**Graph the data (Task 7)**

I decided the best way to start to answer the second research question is 
by creating a scatter plot. I first want to test if there is a positive 
linear relationship between berry weight and TSS in the combined data. Then I 
will test the Bluecrop variety alone to see if cross-pollination affects this 
relationship. I'll do the same with berry weight vs pH and berry weight vs TSS,
but I suspect that there will be a negative linear relationship in both cases. 


```{r warning=FALSE}
#Testing berry weight against TSS using the combined data
ggplot(Data, aes(Berry_weight_g, TSS)) +
  geom_point(size=3, alpha=0.8) + 
  xlab("Berry weight (g)") + ylab("TSS (%)") + 
  ggtitle("Scatterplot of Blueberry Weight (g) vs TSS (%)")

#Testing berry weight against TSS using just the Bluecrop variety
ggplot(Data1, aes(Berry_weight_g, TSS)) + facet_wrap(~Pollen_donor) +
  geom_point(aes(color = Pollen_donor),
  size=3, alpha=0.8) + xlab("Berry weight (g)") + ylab("TSS (%)")

```

When combining the data, it seems like a positive linear relationship may exist 
between these two variables. However, there's no way to know for sure without 
trying to fit the data to a linear model and using regression statistics. The 
relationship is harder to see when separating the data into the different treatments.


```{r warning=FALSE}
#Testing berry weight against pH using the combined data
ggplot(Data, aes(pH, Berry_weight_g)) +
  geom_point(size=3, alpha=0.8) + 
  ylab("Berry weight (g)") + 
  ggtitle("Scatterplot of Blueberry Weight (g) vs pH")

#Testing berry weight against pH using just the Bluecrop variety
ggplot(Data1, aes(pH, Berry_weight_g)) + facet_wrap(~Pollen_donor) +
  geom_point(aes(color = Pollen_donor), 
  size=3, alpha=0.8) + ylab("Berry weight (g)") 

```

Unlike the previous set of figures, it is harder to see a trend in the data, even
before separating the data into the different treatments. I would still try 
fitting the data to a linear model via regression to be sure. 

```{r warning=FALSE}
#Testing berry weight against TA using the combined data
ggplot(Data, aes(Titratable_acid, Berry_weight_g)) +
  geom_point(size=3, alpha=0.8) + 
  ylab("Berry weight (g)") + xlab("Titratable acid (TA)") +
  ggtitle("Scatterplot of Blueberry Weight (g) vs Titratable acid (TA)")

#Testing berry weight against TA using just the Bluecrop variety
ggplot(Data1, aes(Titratable_acid, Berry_weight_g)) + facet_wrap(~Pollen_donor) +
  geom_point(aes(color = Pollen_donor), 
  size=3, alpha=0.8) + ylab("Berry weight (g)") + xlab("Titratable acid (TA)")

```

Similar to the previous set of scatterplots, it's difficult to see a trend. Again,
I'd like to try fitting the data to a linear model.  

**Research question 3**

I suspect that stem diameter may have an effect on berry weight because thicker
stems should carry more water and nutrients, and hold more leaves (relative to 
thinner stems). These attributes could yield larger berries from the additional 
water and nutrients. 

**summarizing the data**

This time, I need to arrange the data to see if there are any apparent trends 
in the stem diameter and the berry weight columns. I also don't need to focus in 
on one variety like I did before because I'm only interested in the possibility 
of stem diameter acting as a confounding variable in my study. If I accidentally
chose thicker stems in some treatments over others, and the relationship I
suspect exists, it could affect my results!

```{r}
Data2<-as_tibble(Pilot) %>%
  select(stem_diam, berry_weight) %>%
  arrange(desc(stem_diam), berry_weight) %>%
  rename(Berry_weight_g = berry_weight) %>%
  rename(Stem_diam_cm = stem_diam)
head(Data2, n=36)
```

There doesn't appear to be any obvious trend organizing the data this way. I'll
have to create a scatterplot to see if a trend becomes more easily visible. 

```{r}
ggplot(Data2, aes(Stem_diam_cm, Berry_weight_g)) +
  geom_point(size=3, alpha=0.8) + xlab("Stem diameter (cm)") +
  ylab("Berry weight (g)") + ggtitle("Scatterplot of Berry weight vs Stem diameter")

```

From the scatterplot above, it is difficult to see if there is a strong relationship
between the berry weight and stem diameter. I would have to try fitting the data
to a linear model via regression to know for sure. 

**Research question 4**

To answer this question, I need to compare the TSS, pH, and TA of the different 
treatments for the Bluecrop variety and for the Duke variety. TSS positively 
correlates with sugar content and sweetness whereas TA positively correlates 
with sourness. A higher quality berry is large, sweet whereas a poor quality 
berry is typically small and sour. On the other hand, berries with a lower pH
are more sour (negative correlation). I've already compared TSS between the 
different treatments of the Bluecrop variety but I still need to compare the 
other variables. I also need to compare the aforementioned variables for the 
different treatments for the Duke variety.

**summarizing the data**

First, I'll organize the data to focus on the three variables mentioned before. 
Then I need to separate the data into the two varieties of interest. 

```{r}
Data3<-as_tibble(Blue) %>%
  select(Variety, Pollen_donor, TSS, pH, Titratable_acid) %>%
  arrange(Variety, Pollen_donor, desc(TSS), pH, Titratable_acid) 

head(Data3, n=36)

Data4<-as_tibble(Data) %>%
  filter(Variety == "Duke") %>%
  select(Variety, Pollen_donor, TSS, pH, Titratable_acid) %>%
  arrange(Variety, Pollen_donor, desc(TSS), pH, Titratable_acid) 

head(Data4, n=36)
```

Before graphing the data, I'll compare the means and medians of the different 
treatments to see if I can see any obvious differences. 

```{r}
#Comparing TSS, pH and TA in the Bluecrop variety
summary(Blue_B$TSS)
summary(Blue_D$TSS)
summary(Blue_R$TSS)

summary(Blue_B$pH)
summary(Blue_D$pH)
summary(Blue_R$pH)

summary(Blue_B$Titratable_acid)
summary(Blue_D$Titratable_acid)
summary(Blue_R$Titratable_acid)
```

There doesn't appear to be a significant difference in TSS or pH between the 
treatments in the Bluecrop variety. However, there may be a significant difference
in the TA between the treatments. I have to test this with an ANOVA to be certain.

Next, I'll organize the data to select for the different treatments of the Duke 
variety. Then I can compare the TSS, pH and TA in the Duke variety. 

```{r}
Duke_D<- as_tibble(Data) %>%
filter(Variety == "Duke") %>%
filter(Pollen_donor == "Duke") 

Duke_B<- as_tibble(Data) %>%
filter(Variety == "Duke") %>%
filter(Pollen_donor =="Bluecrop")

Duke_R<- as_tibble(Data) %>%
filter(Variety == "Duke") %>%
filter(Pollen_donor =="Reka") 

#Comparing TSS, pH and TA in the Duke variety
summary(Duke_D$TSS)
summary(Duke_B$TSS)
summary(Duke_R$TSS)

summary(Duke_D$pH)
summary(Duke_B$pH)
summary(Duke_R$pH)

summary(Duke_D$Titratable_acid)
summary(Duke_B$Titratable_acid)
summary(Duke_R$Titratable_acid)
```

There may be a significant difference between the TSS of the different treatment
groups as well as the TA but there doesn't seem to be much difference in the pH. 
I would need to investigate this further with ANOVA to know for certain (assuming
the data follows a normal distribution).

**Graph the data**

To illustrate the data and look for signs of correlation, I need to create 
boxplots to compare the TSS, pH and TA between the different treatments of the 
two varieties. 

```{r}
ggplot(Blue, aes(Pollen_donor, TSS)) + geom_boxplot() + 
    geom_jitter(width=0.1, alpha=0.5) + xlab("pollen donor") + 
    ylab("total soluble solids (%)") +
    ggtitle("TSS vs Pollen Donor in the Bluecrop variety")

ggplot(Blue, aes(Pollen_donor, pH)) + geom_boxplot() + 
    geom_jitter(width=0.1, alpha=0.5) + xlab("pollen donor") + 
    ylab("pH") +
    ggtitle("pH vs Pollen Donor in the Bluecrop variety")

ggplot(Blue, aes(Pollen_donor, Titratable_acid)) + geom_boxplot() + 
    geom_jitter(width=0.1, alpha=0.5) + xlab("pollen donor") + 
    ylab("titratable acid (mg citric acid/g FW") +
    ggtitle("Titratable acid vs Pollen Donor in the Bluecrop variety")

```

The boxplots help illustrate the range and medians of the data. Again, there
seems like there could be a significant difference between the TSS and TA but 
there's still no way to say for certain without using statistics. 

```{r warning=FALSE}
#Filtering the variety to Duke only 
Du<-as_tibble(Data) %>%
  filter(Variety == "Duke")

#Setting the pollen donor as a factor so I can rearrange for the control first
Du$Pollen_donor <- factor(Du$Pollen_donor, c("Duke", "Bluecrop", "Reka")) 
  
#Creating the boxplots
ggplot(Du, aes(Pollen_donor, TSS)) + geom_boxplot() + 
    geom_jitter(width=0.1, alpha=0.5) + xlab("pollen donor") + 
    ylab("total soluble solids (%)") +
    ggtitle("TSS vs Pollen Donor in the Duke variety")

ggplot(Du, aes(Pollen_donor, pH)) + geom_boxplot() + 
    geom_jitter(width=0.1, alpha=0.5) + xlab("pollen donor") + 
    ylab("pH") +
    ggtitle("pH vs Pollen Donor in the Duke variety")

ggplot(Du, aes(Pollen_donor, Titratable_acid)) + geom_boxplot() + 
    geom_jitter(width=0.1, alpha=0.5) + xlab("pollen donor") + 
    ylab("titratable acid (mg citric acid/g FW") +
    ggtitle("Titratable acid vs Pollen Donor in the Duke variety")
```

From the boxplots above, there seems to be greater differences between some of 
these treatments relative to the Bluecrop variety treatments. There are promising 
signs of a significant difference in TSS between at least two of the treatments. 
There may even be a significant difference in pH between some of the treatments. 
Again, I can't tell for sure without performing statistics. I'll have to check 
the distribution of the data to see if I can use an ANOVA test. 

<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->

While the figures are helpful for visualizing data and for spotting trends, they 
cannot be used alone as evidence for significant differences between population 
means. Below each exploration of the research questions, I explain what types
of statistics I would use. 

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->

The original dataset (Pilot) is definitely untidy. The columns include sample 
IDs that aren't helpful as well as columns used for calculation of fruit set and
titratable acid. The column names are long and unwieldy too. 

Please see lines 87 through 94 for the first example of how I tidied the data to 
select the columns of interest and arrange them in a sensible order. Then I renamed
some of the poorly named columns. 

<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->

```{r}
#The first tibble I created to tidy the data
Data<-as_tibble(Pilot) %>%
  select(Variety, Pollen_donor, berry_weight, fruit_set,
  Average_TSS, Average_pH, TA.true.) %>%
  arrange(Pollen_donor, desc(fruit_set)) %>%
  rename(Berry_weight_g = berry_weight) %>%
  rename(TSS = Average_TSS) %>%
  rename(pH = Average_pH) %>%
  rename(Titratable_acid = TA.true.)

#Referring back to the original dataset to see which columns are missing from the tibble
head(Pilot)

#Untidying the original tibble
Untidy<-as_tibble(Data) %>%
  rename(Average_TSS = TSS) %>%
  rename(Average_pH = pH) %>%
  rename(TA.true. = Titratable_acid) %>%
  arrange(Variety, Pollen_donor)

head(Untidy)

```
Now the "Untidy" data is as close to the original, untidy "Pilot" dataset as I 
can make it (without adding back the missing columns). 

<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelevant columns, etc.). 

<!--------------------------- Start your work below --------------------------->

For milestone 3, I'd like to pursue the first and last research questions.
These research questions focus on my primary hypotheses and predictions rather 
than potential confounding variables. 

As a refresher, these research questions are:

1. Do blueberry flowers of the bluecrop variety experience a greater fruit set 
when they receive pollen from another blueberry variety?

4. Does the Bluecrop variety experience greater benefits from receiving pollen 
from another variety relative to the Duke variety?

<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 
