---
output:
  github_document:
  html_preview: false
---

# Sasha’s Mini Data Analysis Project

Sasha Tuttle October 14, 2021

`{r setup, include=FALSE} knitr::opts_chunk$set(echo = FALSE)`

Loading packages:

``` {r}
#install.packages("devtools")
#devtools::install_github("UBC-MDS/datateachr")
suppressPackageStartupMessages(library(rmarkdown))
suppressPackageStartupMessages(library(devtools))
suppressPackageStartupMessages(library(datateachr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
```

Loading the pilot study data

``` {r}
Pilot<-read.csv("C:/Users/stutt/OneDrive/Desktop/Greenhouse_study/Pilot_study.csv", 
  header=T)
head(Pilot)
Stem<-rename(Pilot, stem_diam_cm = stem_diam)
```

The Pilot_study.csv file contains data from a pilot study where I tested the 
effects of cross-pollination on three different varieties of *V. corymbosum* 
(highbush blueberry). I tested the effects of cross-pollination with regards to 
fruit yield, berry weight, TSS (%), pH, and titratable acid (TA). I hypothesized 
that blueberry flowers that received pollen from a different variety would have 
a higher yield and berry weight relative to flowers that received pollen from the 
same variety. I also hypothesized that blueberry flowers that were 
cross-pollinated would yield higher quality blueberries (high TSS, low pH, low TA)
relative to self-crossed blueberry flowers. Treatments were assigned to different 
branches of the same bush, so all of the measurements were taken for a whole 
cluster of berries. Note that the units for fruit set and TSS are in percent and the units for TA are mg citric acid/g formula weight (FW). 

I am interested in the data in the columns titled: "Variety", "Pollen_donor",
"fruit_set", "berry_weight", "stem_diam", "Average_TSS", "Average_pH", and "TA.true".
The other columns are not of interest as they were used for calculation or are 
simply sample IDs. 

**Task 1: Investigating the datasets in the datateachr package**

``` {r}
dim(Pilot)
typeof(Pilot)
class(Pilot)
```

The pilot study dataset has 36 rows and 15 columns. It is classified as
a data frame.

``` {r}
#?cancer_sample
print(cancer_sample)
typeof(cancer_sample)
dim(cancer_sample)
class(cancer_sample)
```

The cancer_sample data set has 596 rows, 32 columns, and is classified
as a “list”/“data.frame”.

``` {r}
#?parking_meters
print(parking_meters)
typeof(parking_meters)
dim(parking_meters)
class(parking_meters)
```

The parking_meter data set has 10,032 rows, 22 columns, and is
classified as a “list”/“data.frame”.

``` {r}
#?steam_games
print(steam_games)
typeof(steam_games)
dim(steam_games)
class(steam_games)
```

The steam_games dataset has 40833 rows, 21 columns, and is classified as
a “list”/“data.frame”.

Datasets ranked from first to last choice:

1.  Pilot_study
2.  parking_meters
3.  steam_games
4.  cancer_sample

Of the four datasets investigated, my top two choices are the pilot
study and parking meter data. The pilot study interests me because this
dataset needs organizing and I am familiar with it. The parking meter
data set is also ideal because it is smaller relative to the other
datasets provided and it’s easier to understand (relative to the breast
cancer dataset).

**Task 2: Exploring data from the pilot study.**

1.  Plot the distribution of a numeric variable.

I plotted the stem diameter(cm) of the different blueberry varieties to
see if one of the varieties had a greater stem diameter relative to the
other. Thicker stems may have more leaves, allowing branches to produce
a greater number of berries or heavier berries. So, if thicker branches
were accidentally chosen for one variety compared to the other, we may
see a greater yield and berry weight for said variety.

``` {r}
ggplot(Pilot, aes(Variety, stem_diam)) +
  geom_boxplot(width=0.3) + geom_point() + ylab("Stem diameter (cm)")+
  ggtitle("Boxplot of Stem Diameter vs Blueberry Variety ")
```

From the boxplot above, it doesn’t seem like there is a significant
difference between the stem diameter of the two varieties. However, I
should also see if stem diameter differs significantly between different
treatments within the same variety. This could act as a confounding
variable and affect my results.

2.  Explore the relationship between two variables in a plot.

I plotted the stem diameter of the variety against the berry weight to
see if I could find a linear relationship. I hypothesized that clusters
with thicker branches would have more leaves and energy to produce
heavier berries.

``` {r}
ggplot(Pilot, aes(x=stem_diam, y=berry_weight)) +
  geom_point() + ylab("Berry weight (g)") + xlab("Stem diameter (cm)") +
  ggtitle("Scatterplot of Blueberry Weight vs Stem diameter")
```

The plot produced doesn’t suggest a relationship is present between the
two variables. I should analyze the data further with a regression to be
certain.

3.  Filter observations

I filtered my data into two different blueberry varieties (the pollen
recipients in the experiment). Blueberry variety can significantly
affect blueberry weight (among other parameters) and I’m not interested
in this variable. So, I will assess the effect on each variety
separately.

``` {r}
Blue<-filter(Pilot, Variety == "Bluecrop")
head(Blue)
Duke<-filter(Pilot, Variety == "Duke")
head(Duke)
```

4.  Create a tibble

I created a tibble to separate the Bluecrop variety from the Duke
variety. Blueberry variety can have a significant effect on the
variables I measured so I want to assess the effects of
cross-pollination on each variety separately. I arranged pollen_donor
column in descending order so I can more easily compare the self-crossed
treatments against the out-crossed treatments. I also only selected the
fruit set and berry weight columns so I can ignore the columns used in
calculations and for sample IDs.

``` {r}
as_tibble(Pilot) %>%
  select(Variety, Pollen_donor, fruit_set, berry_weight, 
  Average_TSS, Average_pH, TA.true.) %>%
  filter(Variety == "Bluecrop") %>%
  arrange(Pollen_donor, desc(berry_weight)) %>%
  rename(Berry_weight_g = berry_weight) %>%
  rename(TSS = Average_TSS) %>%
  rename(pH = Average_pH) %>%
  rename(Titratable_Acid = TA.true.)
```

**Task 3: Write your research questions**

-   Do blueberry flowers of the Bluecrop variety experience a greater
    fruit set when they receive pollen from another blueberry variety?

-   Do heavier berries have a higher TSS? a lower pH/TA?

-   Does stem diameter have an effect on blueberry weight and/or TSS?

-   Does the Bluecrop variety experience greater benefits from receiving
    pollen from another variety relative to the Duke variety?

**Milestone 2**

